# Fundamentals of Bayesian Inference

The first few chapters of Gelman's text are introductory, and we attempt to highlight the key definitions and summarize each chapter. At the end of each chapter we attempt several problems.  Probability and inference is defined using three steps

1. setting up the full probability model for a joint distribution for all observable and unobservable quantities.
2. Conditioning on observed data: computing the appropriate *posterior* distribution, the conditional probability distribution of the unobserved quantities of oltimate interest, given the observed data.
3. Evaluating the fit of the model.

## General notation for statistical inference

There are two different kinds of estimands, the first are potentially observable quantities, such as future observations of a process, and the second are quantities that are not directly observable, namely the parameters that govern a process being investigated.  

### Exchangeability {-}
One key assumption is that the n values $y_i$ are regarded as *exchangeable*, meaning that the uncertainty can be expressed as a joint probability $p(y_1,...,y_n)$ that is invariant to permutations of indexes.  Often times the exchangeable distribution is modeled as *iid*.

### Explanatory variables {-}
It is common to have observations on each unit which have non-random variables called *explanatory variables* or *covariates*.  The explanatory variables are usually denoted by X.  However treating X as random then exchangeability can be extended $(x,y)_i$ which is invariant to permutations of the indexes.  Further, it is always appropriate to assume exchangeability of y, conditioned on sufficient information of X, where the indexes can be thought of as randomly assigned.  It follows that if two units have the same value of x, then the distributions of y are the same.

### Hierarchical modeling {-}
for a model across patients across different cities, we can assume exchangability to patients within a city.  Further conditioned on the explanatory variables at the individual, the conditional distribution given these explanatory variables would be exchangeable.


## Bayesian inference
The prior, p($\theta$), and the sampling distribution, or the *data distribution*, $p(y|\theta)$ is

$$ p(\theta,y) = p(\theta)p(y | \theta)$$

Where using Bayes' rule the posterior distribution 
\begin{equation}
p(\theta | y ) = \frac{p(\theta)p(y| \theta)}{p(y)}
(\#eq:bayes)
\end{equation}

Where $p(y)  = \int p(\theta)p(y | \theta)d\theta$, or a sum in discrete case.  An equivalent form of \@ref(eq:bayes) is the *unnormalized posterior density* given as

\begin{equation}
p(\theta | y ) \propto p(\theta)p(y | \theta)
(\#eq:unnorm)
\end{equation}

Note that $p(y | \theta)$ is taken as a function of $\theta$, not of y.

## Prediction {-}
Inferences about an unknown *observable* variable, are called predictive inferences.  Beofre the data y are considered, the distribution of the unknown, observable, y is $$ p(y) = \int p(y,\theta)d\theta = \int p(\theta)p(y | \theta)d\theta$$

this is defined as the marginal distribution of y, and also called *prior predictive distribution*.  Prior refers that the data is not conditional on any previous observation, and predictive refers to the data being observable.

The *posterior predictive distribution* is conditional on the observed y, but is predictive because it is predicting observable values.

\begin{equation}
\begin{split}
p(\hat{y} | y) &= \int p(\hat{y}, \theta | y) d\theta \\
&= \int p(\hat{y} |\theta, y) p(\theta|y) d\theta \\
& = \int p(\hat{y}| \theta) p(\theta | y) d\theta \\
\end{split}
(\#eq:postpred)
\end{equation}

## Likelihood {-}
The data *y* affects the posterior inference only through \@ref(eq:unnorm) likelihood function $p(y| \theta)$ which is regarded as a function of $\theta$ for fixed y.  The *likelihood function* is defined as $p(y | \theta)$, and the *likelihood principle* is for any given sample, and any two likelihood models $p(y | \theta)$, two models with the same likelihood will have the same inference for $\theta$.

## Subjectivity and Objectivity {-}

The frequentist models, MLEs, have subjectivity in their assumptions because they rely on long sequence of identical trials, that are iid.  The Bayesian model relies on the prior distribution.  If any experiment is repeatable and can replicated, the the prior distribution can be estimated from the data themselves and the analysis is more 'objective'.  Replication increases objectivity of a given model.  However the Bayesian approach allows for (1) the ability to combine information from multiple sources (allowing for greater objectivity) and (2) more encompassing by accounting for uncertainity about the unknowns in a statistical problem.

It is important to include as much background information as possible 

## Exercises

1.  Suppose for $\theta=1$, then y$\sim N(1,\sigma)$, and if $\theta=2, y\sim N(2,\sigma)$.  Where P($\theta=1)=P(\theta=2)=0.5$.
    (a)  For $\sigma=2$; we must write the formula for the pdf of y.   $p(y) = \sum_\theta p(y|\theta)p(\theta) = (1/2)N(1,\sigma^2)+ (1/2)N(2,\sigma^2)$ as the marginal density.
```{r}
 fy<-function(y) {return(0.5*dnorm(y,mean=1,sd=2)+0.5*dnorm(y,2,sd=2))}    
```

    (b) $P(\theta=1 | y=1) = \frac{p(\theta=1)p(y|\theta=1)}{p(y)}= \frac{(1/2)N(1,4)}{(1/2)N(1,\sigma^2)+ (1/2)N(2,\sigma^2)}$ = 0.53 
    
```{r}
dy<-function(y){ return( (1/2)*dnorm(y,mean=1,sd=2)/fy(y))}
dy(1)
```


4. twelve games with point spread of 8 points.
    (a) Using relative frequency, P(favorite wins $|$ point spread =8) = 0.67.  
    P(favorite wins by at least 8 $|$ point spread =8) = 0.42.  
    and P(fav. wins by at least 8 $|$ spread =8, favorite team wins) = 0.62.
```{r} 
spread<-8

## outcome of the games favor score - underdog score
 games<-c(-7,-5,-3,-3,1,6,7,13,15,16,20,21)
## frequentist approach
 fav.wins<-  mean(games>0)
 message(paste0("(frequentist): fav wins: ", round(fav.wins,2)))
 fav.by.8<- mean((games>8))
  message(paste0("(frequentist): fav wins by 8: ", round(fav.by.8,2)))

  ## P( fav. wins >8 | fav. wins) = P(fav. wins > 8, fav. wins )/ P(fav. wins)
 cond<-sum(games>8 & games>0)/sum(games>0)   
  c<-fav.by.8/fav.wins
 message(paste0("(frequentist): fav wins by 8 given fav. wins: ", round(cond,2)))
```

  (b) now  we assume a normal distribution with $d|x \sim N(-1.25, 10.10)$.  So $P(d > -x)= P(Z\sigma+\mu > -x)=P(Z > -x-\mu / \sigma)$
  (i) Probablity fav team wins is 0.75
  (ii) fav team wins by 8 (beats the spread) is 0.45, we expect this to be 0.5 (the middle of the normal distribution because we centered on the spread)
  (iii) P(wins by 8 $|$ favorite team wins) = P(favorite team wins $|$ wins by 8)P(wins by 8)/P(favorite team wins) = P(wins by 8)/P(fav. team wins) since the conditional prob. =1 given the favorite team wins.  The prob. that they win by at least 8 is 0.6.
```{r}
## part b
  d<-games-8
  sample.mean <-mean(d)
  sample.sd<-sd(d)
  ## assume  d|x ~ N(0,10.10)
  fav.wins.norm<- 1-pnorm(-8,mean=sample.mean,sd=sample.sd)
   message(paste0("(normal): fav wins: ", round(fav.wins.norm,2)))

  fav.by.8.norm<-1-pnorm(0,mean=sample.mean,sd=sample.sd)
    message(paste0("(normal): fav wins by 8: ", round(fav.by.8.norm,2)))
## Pr(Wins by 8 | Fav. wins) = P(Fav. wins | wins by 8)P(wins by 8) / P(fav. wins)
    ## P(Fav. wins | wins by 8) = 1
  cond.norm<-fav.by.8.norm/fav.wins.norm
   message(paste0("(normal): fav wins by 8 given fav. wins: ", round(cond.norm,2)))

```
5. We need to estimate the probability that there is at least one congressional election that is tied in the next U.S. election.  There are 435 senate elections.
    (a)  The parameters of interest are $\theta_i$ the true probability that the election is tied.  We can let the *prior* $\theta \sim  Beta(\alpha,\beta)$.  The *likelihood* is $y|\theta_i \sim Binomial(435,\theta_i) = \theta^{\sum y_i}(1-\theta)^{435-\sum y_i}$ follows a Binomial distribution (ignoring the binomial coefficient) where we assume each election is independent. Hence the posterior for theta $f(\theta | y) \sim Beta(\sum y_i +\alpha, n-\sum y_i +\beta)$. where $\alpha,\beta$ are set to 1 for the uniform prior.  For this case we set $\alpha, \beta$ equal to 1, 10 which has a prior mean of 0.09.
    
```{r}
theta=seq(from=0,to=1,by=.01)
plot(theta,dbeta(theta,1,10),type='l')
```

(b)  In the period of 1900-1992, there were 20,597 elections, out of which 6 were decided by less than 10 votes,  and 49 were decided by less than 100 votes.  
      we can estimate the probability of a tie to be less than 6/20,597 and bounded by 49/20,597.  So for the Binomial trials the sum of the successes is 6, and n=20,597, so the posterior could be $\theta | y \sim Beta(1+6, 10+20,597 -6)$ is the posterior for $\theta$.  This assumes that 10 votes is within the neighborhood of an election tie.   
      The question asks to compute at least one election tie, from a total of 435 elections.  This follows a Binomial(435, $\hat{\theta})$. Where we use the posterior mean to estimate $\theta$.  The posterior mean using the Beta(7,20601) yields a mean of $\hat{\theta}=\frac{7}{20608} = 3.4e-04$ as the posterior mean.
      
      
  Then the probability that at least 1 election is tied, from 435 total elections will  follow a Binomial(435, $\hat{\theta})$, where we can use the posterior distribution for $\theta | y$ in the Binomial likelihood $P(X\geq 1 | \hat{\theta})= 1-P(X\leq0 | \hat{\theta})$  which has a probability of 0.14 of at least 1 election tie.
```{r}
 # the posterior for theta is Beta(1+6,10+20597-6)
 plot(theta,dbeta(theta,7,10+20597-6),type='l')

 ## posterior mean is 7/(20601)
## then P(X>=1) = 1-P(X<=0 | p)
  1-pbinom(0,435,prob=7/20608)
```

9. A clinic has three doctors.  Patients come into the clinic at random, starting at 9 a.m. according to a Poisson process, with a time parameter, t, of 10 minutes; that is after opening the first patient appears follows an exponential distribution with average waiting time of 10 minutes. Then the next patient arrives with a waiting time of an expected 10 minutes as iid exponential distribution.  After a patient arrives, the patient waits until a doctor is available, and the doctor visits a patitient uniformly between 5-20 minutes.  The clinic stops admitting patients at 4 pm, and closes after the last patient is completed with the visit.

   (a) Simulate this process once. how many patients visited the office? how many had to wait for a doctor? what was the average wait? when did office close?
   
```{r}

 
 
### waiting time for a new patient to arrive in the clinic
###########################################################
 # patientList is the data frame of all patients
 # closeTime is the time to stop admitting (420 minutes)
 # currentPatient Number 
 # current time is the running total of time
  newPatientArrival<-function(patientList,
                             closeTime=timeToClose,
                              waitTime,
                             visitTime,
                             currentPatientNumber=0,
                             currentTime,
                             assignedDoctor="none",
                             completionTime=0){
    # waiting time for next patient
     patientTime<-round(rexp(1,rate=1/10),2)
    # current time of existing patients
     current<-max(patientList$currentTime)
   ## the clinic stops admitting patients at 4pm
   if( (current+patientTime)<=closeTime){
     ## in minutes
   newPatient<-createPatientChart(currentPatientNumber,patientTime,waitTime,visitTime,currentTime,assignedDoctor,0)
    }else{
    newPatient<-createPatientChart(currentPatientNumber,patientTime,waitTime,visitTime,currentTime,"closed_notAdmitted",0)
   }
   return(newPatient)
 }
#################### 
 
 
 
 computeWaitTime<-function(doctors=NULL,
                           patientList=NULL,
                           patientID=1){
   ## need to compute visiting time (booked)
   ## next time available
   ## required input current time for a specific doctor/patient ?
   # patient time (minutes)
   
   ## FIX ME: it is grabbing 2 patient IDs?
   currentTime<-patientList$currentTime[which(patientList$patient==patientID)]
   visitTime<-runif(1,min=5,max=20) ## minutes
 
   
   if(any(doctors$nextTimeAvail<currentTime)){
     waitTime=0
     assignedDr<-sample(doctors$dr[which(doctors$nextTimeAvail<currentTime)],1)
     ### current time + visitTime
     nextAvailTime<- visitTime+currentTime+waitTime
     ## completion time for patient exit (closing time).
   }else if(any(doctors$nextTimeAvail<currentTime)==FALSE){
     # all doctors are booked, no available doctors.
     # wait time is the difference between next available time (assuming all times are greater than patient time)
     waitTime<-min(doctors$nextTimeAvail-currentTime)
     assignedDr<-doctors$dr[which( (doctors$nextTimeAvail-currentTime)==min(doctors$nextTimeAvail-currentTime))]
       if(length(assignedDr)>1){
         assignedDr<-assignedDr[1]
       }
     nextAvailTime<- visitTime+currentTime+waitTime  ## completion time for patient to exit
    }## if all doctors unavail
    #print(assignedDr)
    #print(currentTime)
    ##  update doctor list
     doctors[which(doctors$dr==assignedDr),'visitingPatient']<-patientID
     doctors[which(doctors$dr==assignedDr),'nextTimeAvail']<-nextAvailTime
     doctors[which(doctors$dr==assignedDr),'currentTime']<-currentTime ## patient time
     doctors[which(doctors$dr==assignedDr),'visitTimeLength']<-visitTime
     # flag avail to no.
     doctors[which(doctors$dr==assignedDr),'avail']<-'no'
     ## update patient list
     patientList[which(patientList$patient==patientID),'doctorWaitTime']<-waitTime
     patientList[which(patientList$patient==patientID),'doctorVisitTime']<-visitTime
     patientList[which(patientList$patient==patientID),'assignedDoctor']<-assignedDr
     patientList[which(patientList$patient==patientID),'completionTime']<-nextAvailTime
   return(list(patient=patientList,doctor=doctors))
 }
 
 
 ## creates a patient object
 createPatientChart<-function(currentPatientNumber,arrivalTime,waitTime,visitTime,currentTime,assignedDoctor,completionTime){
   patientID<-data.frame(patient=currentPatientNumber+1,
                         arrivalTime=arrivalTime,
                         doctorWaitTime=waitTime,
                         doctorVisitTime=visitTime,
                         currentTime=currentTime,
                         assignedDoctor=assignedDoctor,
                         completionTime=0)
   return(patientID)
 }
 
 updatePatientList<-function(patientList,patientID){
   patientList<-rbind(patientList,patientID)
   return(patientList)
 }
 
updateTime<-function(currentTime,newTime=NULL,p1){
  p1$currentTime<-currentTime+newTime
  return(p1)
}
 totalPatients<-0
## this is the simulation
 ## first task : loop through the time update for patients
 ## second task : include the doctor assignment query.
simulateProcess<-function(doctors=NULL,
                          totalWait=NULL,
                          totalPatients=0,
                          timeToClose=420,
                          currentTime=NULL){
  ## initiate Patient List
 patientList<-data.frame(patient=0,
                         arrivalTime=0,
                         doctorWaitTime=0,
                         doctorVisitTime=0,
                         currentTime=0,
                         assignedDoctor='none',
                         completionTime=0)
 ## not sure what to put here.
 currentTime<-patientList$currentTime[which(patientList$patient==max(patientList$patient))] ## current time is the max current time from patient chart.
 currentPatientNumber<-0
 
 ## timeToClose (minutes) is stopping to admit patients
  while(currentTime<timeToClose){
    ## patient enters after the (i-1) patient enters.
    p1<-newPatientArrival(patientList,
                             closeTime=timeToClose,
                          waitTime=0,
                          visitTime=0,
                             currentPatientNumber=currentPatientNumber,
                             currentTime)
    ## update time
    p1<-updateTime(p1$currentTime,newTime=p1$arrivalTime,p1)
    
    # given a patient time, switch the availability of any doctor
    # if a doctors next available time is less than the current time, switch him to available
    ## FIX ME: need to ensure this flag is correct.
    if(any(doctors$nextTimeAvail<p1$currentTime)){
      doctors$avail[which(doctors$nextTimeAvail<p1$currentTime)]<-'yes'
    }
    
    ## create a patient list
    if(currentPatientNumber==0){
      patientList<-p1
     # update patient number  
      currentPatientNumber<-currentPatientNumber+1
    }else{
      patientList<-rbind(patientList,p1)
     # update patient number  
      currentPatientNumber<-currentPatientNumber+1
    }
    
    ## task 2 assign a doctor
     ### check for doctor availability
     ## compute wait time, and/or compute the next available time
     ## returns a list object.
    clinicList<-computeWaitTime(doctors,patientList,patientID=patientList$patient[currentPatientNumber])
  
    doctors<-clinicList[["doctor"]]
   patientList<-clinicList[["patient"]]
   ## update flags
     # update currentTime
    ## current time is cumulative sum of the arrival times.
   currentTime<-patientList$currentTime[which(patientList$patient==max(patientList$patient))] ## current time is the max current time from patient chart.
   
   ## fix me:
   ## reset doctor availability based on current patient time.
   upID<-which(doctors$nextTimeAvail<currentTime)
   doctors$nextTimeAvail[upID]<-currentTime
   doctors$currentTime[upID]<-currentTime 
   doctors$visitTimeLength[upID]<-0 
  }## while loop
  return(list(patient=patientList,doctors=doctors))
}

```

```{r,eval=TRUE}

 doctors<-data.frame(dr=c('a','b','c'),
                     visitingPatient=c(0,0,0), ## who is doctor seeing (patient ID)
                     visitTimeLength=c(0,0,0), # length of doctor visit U(5,20)
                     currentTime=c(0,0,0),    ## current Time
                     nextTimeAvail=c(0,0,0),  ## current time + visitTimeLength = next avail time.
                     avail=c("yes","yes","yes"))
## initiate times
 totalWait<-0
 currentPatientNumber<-0
 ## clinic opens at 9am -4pm that is 7 hours (420 min.)
 timeToClose<-7*60 ## stops admiting patienets in 420 minutes
 ## current time is 0
 ## this will be the running total of minutes.
 currentTime<-0
 
 
 
res<-simulateProcess(doctors,
                          totalWait,
                          totalPatients,
                          timeToClose,
                          currentTime)

pl<-res$patient[which(res$patient$currentTime<=420),]

 hist(pl$doctorVisitTime)
 abline(v=(20+5)/2) ## should be ~12
 
 hist(pl$arrivalTime) ## should be close to 10  exp(1/10) has mean 10
 
  hist(pl$doctorWaitTime[which(pl$doctorWaitTime!=0)]) ## about 2.41
  print(max(pl$completionTime)-420) ## closing time
  print(max(pl$patient)) ## total patient  should be 42

## (20-5)/6 + 10 this is about 12.5 minutes of arrival + visit time. which is approximately close to 10 
## the arrival time is about 10 minutes.  

## we should expect 42 patients
 #420/10
  
  ## sanity check
  #all(pl$currentTime+pl$doctorWaitTime+pl$doctorVisitTime-pl$completionTime==0)
```

## Simulation 100 times {-}

total number of patients was approximately 42, which we expect since the total 420/10.  The total number waiting with 3 doctors is 6.61 for 1 day. the average waiting time was about 4-5 minutes.  For 1 day, the average closing time was 5.32 minutes after 4 pm  
 
 

```{r,simul,eval=T}

 totalPat<-NULL
 totalWaiting<-NULL
 avgWaiting<-NULL
 closing<-NULL
 patientList<-NULL
 p1<-NULL
 
for(i in 1:100){
  
 doctors<-data.frame(dr=c('a','b','c'),
                     visitingPatient=c(0,0,0), ## who is doctor seeing (patient ID)
                     visitTimeLength=c(0,0,0), # length of doctor visit U(5,20)
                     currentTime=c(0,0,0),    ## current Time
                     nextTimeAvail=c(0,0,0),  ## current time + visitTimeLength = next avail time.
                     avail=c("yes","yes","yes"))
## initiate times
 totalWait<-0
 totalPatients<-0
 currentPatientNumber<-0
 ## clinic opens at 9am -4pm that is 7 hours (420 min.)
 timeToClose<-7*60 ## stops admiting patienets in 420 minutes
 ## current time is 0
 ## this will be the running total of minutes.
 currentTime<-0
 
 
res<-simulateProcess(doctors,
                          totalWait,
                          totalPatients,
                          timeToClose,
                          currentTime)

pl<-res$patient[which(res$patient$currentTime<=420),]


 totalPat<-c(totalPat,max(pl$patient))
  totalWaiting<-c(totalWaiting,nrow(pl[which(pl$doctorWaitTime!=0),]))
 avgWaiting<-c(avgWaiting,mean(pl[which(pl$doctorWaitTime!=0),"doctorWaitTime"]))
 closing<-c(closing,max(pl$completionTime))
 
}
 
 hist(totalPat,main="total patients")
 abline(v=420/10,col='red')
 
 hist(totalWaiting,main="total number waiting")
 hist(avgWaiting,main="avg. waiting (min)")
 hist(closing-420,main="closing time")

```


