# Single parameter models {#cross}

## Estimating a probability from binomial data

\begin{equation}
p(y  | \theta) = {n \choose y}\theta^y(1-\theta)^{n-y}
(\#eq:binomialProb)
\end{equation}

To perform Bayesian inference we assume $\theta \sim U(0,1)$ where the posterior is 

\begin{equation}
p(\theta | y) \propto \theta^y(1-\theta)^{n-y}
(\#eq:binomialPosterior)
\end{equation}

which is the form of a `beta` distribution $\theta | y \sim Beta(y+1, n-y+1)$

## Posterior as a compromise between data and prior information
 The posterior is less variable than the prior because it incorporates the information from the data.
 
\begin{equation}
E(\theta) = E(E(\theta | y))
(\#eq:adamsLaw)
\end{equation}

\begin{equation}
V(\theta) = E(V(\theta|y)) + V(E(\theta | y))
(\#eq:evesLaw)
\end{equation}
 
where $\theta|y$ is the posterior.  So the average of the prior, is the average of the posterior means over the distribution of possible data.  The variance of the prior \@ref(eq:evesLaw) says the posterior variance is on average smaller than the prior variance.

## Summarizing the posterior inference
The mean, median, mode, and standard deviation of the posterior distribution summarize the all the current information about a model.  

### Posterior quantiles and intervals{-}

The posterior uncertainty can be reported by presenting the quantiles of the posterior distribution.  The interval, a *central interval of posterior probability* corresponds to the case of 100($1-\alpha)\%$, to the range of values above and below which lies exactly 100($\alpha/2)\%$ of the posterior probability.  The interval estimates are *posterior intervals*.  This differences from the confidence interval because the confidence interval is not a probability interval, because either the parameter is within the region or it is not, but the confidence interval provides information in the long run over repeated experimentation as to how many experiments would contain the true parameter.

There is also the *highest posterior interval* which is a probabilistic interval that is not less than any region outside of the interval.

## Informative prior distributions

the property that the posterior distribution follows the same parametric form as the prior distribution is called *conjugacy*.  Where the beta prior distribution is a *conjugate family* for the binomial likelihood.  

so given the binomial likelihood $p(y|\theta)\propto \theta^a(1-\theta)^b$, and a prior density $p(\theta)\propto \theta^{\alpha-1}(1-\theta)^{\beta-1}$ the posterior is of the beta family.


$$
\begin{aligned}
 p(\theta | y) &\propto \theta^y(1-\theta)^{n-y}\theta^{\alpha-1}(1-\theta)^{\beta-1}\\
 &= \theta^{y+\alpha-1}(1-\theta)^{n-y+\beta-1}\\
 &= Beta(\theta | \alpha+y, \beta+n-y)
\end{aligned}
$$


### Conjugate prior distributions{-}
Conjugacy is formally defined as if F is a class of sampling distributions $p(y|\theta)$, and P is a class of prior distributions for $\theta$, then the class P is conjugate for F if $p(\theta | y) \in P$ for all $p(.|\theta)\in F$ and $p(.)\in P$.

### Conjugate prior, distributions, exponential families, and sufficient statistics{-}
Posterior distributions can be derived using sufficient statistics from exponential families.  The exponential family is defined as 

$$
\begin{aligned}
 p(y_i | \theta)=f(y_i)g(\theta)e^{\phi(\theta)^Tu(y_i)}
\end{aligned}
$$
Where $\phi(\theta), u(y_i)$ are vectors of equal dimension to that of $\theta$.  The $\phi(\theta)$ is called the *natural parameter* for the family (F).  The likelihood of a sequence $y=(y_1,...,y_n)$ iid is

$$
\begin{aligned}
 p(y | \theta)&= (\prod_{i=1}^n f(y_i)) g(\theta)^n exp(\phi(\theta)^T \sum_{i=1}^n u(y_i)) \\
  &\propto g(\theta)^ne^{\phi(\theta)^Tt(y)}, t(y)=\sum_{y=1}^n u(y_i)
\end{aligned}
$$

The *sufficient statistic* for $\theta$ is $t(y)$ because the likelihood for $\theta depends on the data, y, only through the value of t(y).

Sufficient statistics benefit posterior distributions because if the prior density is specified as 
$$
\begin{aligned}
 p(\theta)&\propto g(\theta)^{\eta} e^{\phi(\theta)^T \nu} \\
 \end{aligned}
$$
Then the posterior density using sufficient statistics is

$$
\begin{aligned}
 p(\theta | y ) &\propto g(\theta)^{\eta+n} e^{\phi(\theta)^T (\nu+t(y))} \\
 \end{aligned}
$$

Exponential families are the only classes of distributions that have natural conjugate prior distributions. 

## Normal distribution with known variance

The normal distribution is foundational to statistical modeling, with the central limit theorem (CLT) allowing for the use of normal likelihood in many statistical problems which can approximate complex likelihoods.  If the normal distribution does not provide a good model fit, finite mixtures of distributions can identify useful solutions.

### Likelihood of one data point{-}

With mean $\theta$ and known variance $\sigma^2$ the sampling distribution of a given point is defined 

$p(y|\theta) = \frac{1}{\sqrt{2*\pi*\sigma^2}}e^{-\frac{1}{2\sigma^2} (y-\theta)^2}$

### Conjugate prior and posterior distributions
The prior has the exponential family form given as $\theta \sim N(\mu_0, \tau_0^2)$


$$
\begin{aligned}
 p(\theta ) &\propto exp( \frac{1}{2\tau_0^2}(\theta-\mu_0)^2) \\
 \end{aligned}
$$

WHere completing the square can find the posterior distribution

$$
\begin{aligned}
 p(\theta | y ) &\propto exp(-1/2(\frac{(y-\theta)^2}{\sigma^2} +\frac{(\theta-\mu_0)^2}{\tau_0^2} )  ) \\
 p(\theta | y ) &\propto exp( \frac{1}{2\tau_1^2}(\theta-\mu_1)^2),  \theta|y \sim N(\mu_1,\tau_1^2)
 \end{aligned}
$$
where $\mu_1 = \frac{1/\tau_0^2 \mu_0 + 1/\sigma^2 y}{1/\tau_0^2 + 1/\sigma^2}$ and $1/\tau_1^2 = 1/\tau_0^2 +1/\sigma^2$

In manipulating the distributions the inverse of the variance is defined as the *precision*.  The posterior precision is equal to the prior precision plus the data precision.  And the posterior mean is a weighted average of the prior mean and the observed value, y, proportional to the total precision.

### Posterior predictive distribution{-}

the posterior predictive distribution of a future observation, x, p(x|y) can be calculated 

$$
\begin{aligned}
p(x|y) &= \int p(x|\theta)p(\theta|y) d\theta
&\propto \int exp(-1/2\sigma^2 (x-\theta)^2) exp(-1/2\tau_1^2 (\theta-\mu_1)^2) d\theta
\end{aligned}
$$
 the future observations, x, does not depend on the past observations y given $\theta$.  


### Normal model with multiple observations{-}

For multiple observations, y, the posterior density is formulated as:

$$
\begin{aligned}
 p(\theta | y ) &\propto p(\theta)p(y|\theta) \\
   &=p(\theta)\prod_i p(y_i | \theta)\\
   &\propto exp(-1/2\tau_0^2 (\theta-\mu_0)^2)\prod_i exp(-1/2\sigma^2 (y_i -\theta)^2)\\
   &\propto exp(-1/2 (1/\tau_0^2(\theta-\mu_0)^2+ 1/\sigma^2\sum_i(y_i-\theta)^2) )\\
 \end{aligned}
$$
Simplfying the algebra shows the posterior depends on y only through the sample mean (sufficient statistic), $\bar{y}$ is the sufficient statistic for $\theta$, and the final model is $\bar{y} | \theta,\sigma^2 \sim N(\theta, \sigma^2/n)$

## Other standard single-parameter models


## Exercises

### Question 1{-}
prior Beta(4,4), where a coin is tossed 10 times and heads appears fewer than 3 times.  the exact posterior is Beta(4+y, 4+10-y) for y=0,1,2.  Since we don't know the observed heads, but that $y<3$ we plot the posterior distributions for each possibility.  For 2 heads it is closer to the prior, with posterior mean of 0.33,  which is closest to the prior mean of 1/2.
```{r}
 theta<-seq(from=0,to=1,by=0.01)

 plot(theta,dbeta(theta,4,4),type='l',ylim=c(0,5)) ## prior
 lines(theta,dbeta(theta,4+1,4+10-1),lty=2) ## 1 success
   lines(theta,dbeta(theta,4+2,4+10-2),lty=3,col='red') ## 2 succe
   lines(theta,dbeta(theta,4,4+10),lty=4,col='blue')  ## 0 successes
  legend(0.6, 4, 
         legend=c("prior", "0 H", "1 H", "2 H"),
       col=c("black","blue","black", "red"), lty=c(1,4,3,2), cex=0.8)
```


### Normal approximation example{-}


For female births we have beta(438,544) we use the normal approximation.  This replicates Gelman's Figure 2.3 (a, b) 
```{r}
  theta<-seq(from=0.001,to=1,by=0.01)
  ## example births
   postMean <-function(alpha,beta,y,n){
     return( (alpha+y)/(alpha+beta+n))
   } 
   postVar<-function(alpha,beta,y,n){
     return( ((alpha+y)*(beta+n-y))/( (alpha+beta+n)^2*(alpha+beta+n+1)) )
   }
   sdnorm<-sqrt(postVar(438,544,0,0))
   logitMean<-log( postMean(438,544,0,0)/(1-postMean(438,544,0,0)))
   
  logitTheta<-log(theta/(1-theta))

  par(mfrow=c(1,2))
    plot(theta,dbeta(theta,438,544),type='l',xlim=c(0.35,0.55),main="posterior beta") ## prior
    abline(v=0.446,col='red')
  draws<-rbeta(1000,438,544)
  plot(logitTheta,dnorm(logitTheta,mean =logitMean, sd=sd(log(draws/(1-draws))) ),type='l', xlim=c(-0.5,0.1),main="Normal approx.") ## prior
    abline(v=-0.22,col='red')

```



### Question 3{-}


The prior predictive distributions for the number of 6's in a fair roll, tossed 1,000 times will follow a beta distribution.  Let y be the number of 6's in 1000 rolls of fair die, the probability for a 6 is 1/6, so the number of 6's in this trial is approximatley 167, and 833 failures as the prior prediction.  

We plot the beta distribution of the expected number of heads in 1000 tosses.  

The normal approximation shows the probability of heads in a given 1000  tosses, using a non-informative prior beta(167,833) which has a prior predictive mean of $exp(-1.79)/(1+exp(-1.79)) = 0.143$.  This is not the same for number of success in 1000 tosses.  

We find the probability distribution of a given success and the prior probability predictive interval follows a beta with 95% (0.12,0.17) for hte probability of rolling a 6
```{r}
  ## based on normal approximation sketch the distribution of y
  ## for normal we use the logit transform
  ## n = 1,000
  ## first lets construct a beta distribution.
  ## let the prior be beta(4,4) or even beta(1,1)
par(mfrow=c(1,2))

   plot(theta,dbeta(theta,167,833),type='l',xlim=c(0.1,0.5),main="prior beta") ## prior
  # plot(theta,dbeta(theta,1,5),type='l',xlim=c(0.1,0.5), main='prior beta') 
   abline(v=0.167,col='red') ## 
   #  qbeta(c(0.05,0.25,0.5,0.75,0.95),167,833)
    ## predictive on normal scale.
     draws<-rbeta(1000,167,833)
     logitMean<-log( 167/1000)
  sdd<-sd(log(draws/(1-draws)))

   
  plot(logitTheta,dnorm(logitTheta,mean =logitMean, sd=sdd ),type='l',main="Normal approx.of a total heads", xlim=c(-2,-1.4)) ## prior
    abline(v=logitMean,col='red')
    ### the number of heads.    
  message("prior predictive mean:" ,round( exp(logitMean)/(1+exp(logitMean)),2))

 qx<-(qnorm(c(0.05,0.25,0.5,0.75,0.95),mean =logitMean, sd=sdd ))
 ## this is the probability of landing 6 for theta
 exp(qx)/(1+exp(qx))
## wald 95% prior interval
c( exp(logitMean-1.96*sdd)/(1+exp(logitMean-1.96*sdd)),exp(logitMean+1.96*sdd)/(1+exp(logitMean+1.96*sdd)))
```
### Question 4{-}
We have a mixture of 3 normal distributions, and show the central intervals for 5,25,50,75, and 95$\%$ predictive probabilities.  The question gives $\theta$ as the probability of a 6 on a die, possibly unfair, in 1,000 tosses.  we have $\theta= 1/12,1/6, 1/4$ types of biased die. Using the normal approximation, the predictive prior probability is $\sum_{\theta}p(\theta)p(y| \theta)$ where the likelihood is approximated using a normal distribution $\mu = n*\theta_i, \sigma = n*\theta_i(1-\theta_i)$
```{r}
x<-seq(0,1000)-0.5 # continuity correction.
theta<-c(1/12,1/6,1/4)
n=1000
normLikelihood<-function(x,n,p){
  mean<-n*p
  varz<-n*p*(1-p)
  return(dnorm(x,mean=mean,sd=sqrt(varz)))
  
}
a<-dnorm(x,mean=n*theta[1],sd=sqrt(n*theta[1]*(1-theta[1]))  )
b<-dnorm(x,mean=n*theta[2],sd=sqrt(n*theta[2]*(1-theta[2]))  )
c<-dnorm(x,mean=n*theta[3],sd=sqrt(n*theta[3]*(1-theta[3]))  )


mypri<-a*0.25+b*0.5+c*0.25

sum(mypri) ## sums to 1 it is a distribution

 par(mfrow=c(3,2))
plot(x,mypri,type='l',main='predictive prior number of heads')



data<-data.frame(x=x,p=mypri)

## highest probability interval
# 95%

plot(x,mypri,type='l',main=' 95% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.025),1])
)
abline(v=max(data[which(1-cumsum(data$p)>0.025),1]))

plot(x,mypri,type='l',main='75% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.125),1]))
abline(v=max(data[which(1-cumsum(data$p)>0.125),1])
)

plot(x,mypri,type='l',main='50% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.25),1]))
abline(v=max(data[which(1-cumsum(data$p)>0.25),1])
)

plot(x,mypri,type='l',main='25% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.375),1]))
abline(v=max(data[which(1-cumsum(data$p)>0.375),1])
)

plot(x,mypri,type='l',main='5% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.475),1]))
abline(v=max(data[which(1-cumsum(data$p)>0.475),1])
)
```


### Question 8 (Normal distribution with unknown mean){-}
A random sample of n students is drawn from a large population, and weights are measured.  The average height of the n sampled students is $\bar{y}= 150$ lbs.  Assume the weights in the population are normally distribution with unknown mean, $\theta$, and known standard deviation 20 lbs.  Suppose the prior for $\theta \sim N(180, 40^2)$

For known variance, the limit of the posterior is $p(\theta,y)\approx N(\theta| \bar{y}, \sigma^2/n)$.  And the direct formulation is $p(\theta | \bar{y})= N(\theta | \mu_n,\tau_n^2).$ using equations 2.12.

For a posterior predictive interval, the marginal distribution for new data $p(\tilde{y}  | y) \sim N(\mu_n, \sigma^2+\tau_n^2)$
```{r}
mu_n<-function(mu0,ybar,n,tau02,sigma2){
  mun<- (mu0/tau02 + (ybar*n)/sigma2)/(1/tau02 + n/sigma2)
  return(mun)
}
taun2<-function(tau02,n,sigma2){
  inv.taun2<- 1/tau02 +n/sigma2
  return(1/inv.taun2)
}

ybar=150
sigma2=20^2
mu0=180
  tau02=40^2
  n=10

## posterior interval n=10
  #lower<-round(qnorm(0.025,mu_n(mu0,ybar,tau02,10,sigma2),sd=sqrt(taun2(tau02,10,sigma2))),2)
    upper<-round(qnorm(c(0.025,0.975),mu_n(mu0,ybar,tau02,10,sigma2),sd=sqrt(taun2(tau02,10,sigma2))),2)

message("posterior interval n=10: ",upper[1]," ",upper[2])

## posterior predictive interval
  upper2<-round(qnorm(c(0.025,0.975),mu_n(mu0,ybar,tau02,10,sigma2),sd=sqrt(taun2(tau02,10,sigma2)+sigma2) ),2)

message("posterior predictive interval n=10: ",upper2[1]," ",upper2[2])


 lower<-round(qnorm(0.025,mu_n(mu0,ybar,tau02,100,sigma2),sd=sqrt(taun2(tau02,100,sigma2))),2)
    upper<-round(qnorm(0.975,mu_n(mu0,ybar,tau02,100,sigma2),sd=sqrt(taun2(tau02,100,sigma2))),2)

message("posterior interval n=100: ",lower," ",upper)
 ## this approximately equals the limit
print("The asymptotic approximation")
 qnorm(c(0.025,0.975),mean=150, sigma2/100)
## posterior predictive interval
 lower2<-round(qnorm(0.025,mu_n(mu0,ybar,tau02,100,sigma2),sd=sqrt(taun2(tau02,100,sigma2)+sigma2) ),2)
    upper2<-round(qnorm(0.975,mu_n(mu0,ybar,tau02,100,sigma2),sd=sqrt(taun2(tau02,100,sigma2)+sigma2) ),2)

message("posterior predictive interval n=100: ",lower2," ",upper2)
```


### Question 10{-}
suppose y1,...,y5 are iid Cauchy($\theta,1$) r.vs. and the prior distribution for$\theta \sim U[0,100]$.  the given observations are y=43,44,45,46.5,47.5.

- (a) compute the unnormalized posterior density on a grid of points $\theta=0 1/m, 2/m, ... 100$.  using the grid approximation, compute and plot the posterior density as a function of $\theta$
```{r}
 y= c(43,44,45,46.5,47.5)
 m=1000
theta<-seq(from=0,to=100000)/m
 ## p(theta | y) ~ p(y|theta)*p(theta)
 unnorm.post<-dcauchy(y,location=theta,scale=1)*dunif(theta,min=0,max=100) ## un norm post
 
 post<-unnorm.post/sum(unnorm.post)
 plot(theta,post,type='l',main='Normalized posterior')
```

- (b) Sample 1000 draws from theta from posterior density and plot histogram
we sample from theta [0,100] using the grid approximation, and the probability is from the posterior distribution.
```{r}
  samps<-(sample(theta,1000,prob=post,replace=T))
  hist(samps,main=mean(samps))
  ##sample mean is close to the mean of the observed.

```
- (c) Using the previous 1000 samples of $\theta$ to obtain 1000 samples from the predictive distribution of a future observation $y_6$, and plot the predictive draws.
we use the sampled thetas from the posterior to sample from the Cauchy distribution.  The predictive probability follows $p(x|y) =\int p(x|\theta)p(\theta|y)d\theta$ where $p(x|\theta)$ follows from the Cauchy distribution, and we have the posterior values for each theta (given the uniform grid of thetas).  then for each predictive value, we sum the total probability across all thetas to compute the predictive probability.  The maximum predictive value probability 44.745 with probability of 0.10.
```{r}
## predictive distribution ? ?? 
 # p(x | y) = int p(x|theta)*p(theta|y ) dtheta
## the posterior is p(theta|y) 
 # the likelihood p(x|theta)  ## we use the sampled thetas using the posterior
 ytilde<-rcauchy(1000,location=samps,scale=1) 
 ## probability of the samples 
 #prob_samp<-post[match(samps,theta)]
  summary(ytilde) ## we have a wide distribution of predictive values.
  
## for all predictive values,  find the total probability  
   ## int p(x|theta)*p(theta|y) d\theta  note that p(x|theta) is a function of theta, so we input hte theta grid.
 pred.prob<-sapply(ytilde,function(x) sum(dcauchy(x,location=theta,scale=1)*post))  
  
  plot(ytilde,pred.prob, main='predictive distribution')
  
  ## maximum predictive probability
  message('max pred. prob ', round(ytilde[which(pred.prob==max(pred.prob))],3))
```

### Question 13{-}
- (a) Use the normal approximation to gamma and poisson to determine a posterior for fatal accidents using table 2.2.  Compute the 95$\%$ predictive interval.

We set the prior alpha =25 and beta =1 because the effective sample size ($\beta$) is mean is approximately 25, and given these the quantiles of the prior distribution is 16.2 and 35.7 which contains the observed data well. and the prior mean is approximately 25 which is reasonably close to the table 2.2 values, although a sensitivity analysis is recommended.

We use the conjugate prior for Gamma to find the posterior distribution, which closely is approximated by the normal distribution.  The posterior distributions are similar.   The normal approximation prior followed $N(\alpha/\beta, \sqrt{\alpha}/\beta)$.  we set the variance to be known $\sigma^2=100$.   Using the gamma conjugate prior the predictive interval is $95\% $ (21.1,26.9).



```{r}
# data entry
 year<-seq(1976,1985)
 fatal<-c(24,25,31,31,22,21,26,20,16,22)
 death<-c(734,516,754,877,814,362,764,809,223,1066)
 rate<-c(0.19,0.12,0.15,0.16,0.14,0.06,0.13,0.13,0.03,0.15)
 data<-data.frame(year=year,fatal=fatal,pass.deaths=death,death.rate=rate)
 
 ## gamma prior on theta>0
 theta=seq(from=0,to=50,by=.01)

 ## normal approximation to gamma
  ## Gamma(a,b) ,  mean = a/b,  var = a/b^2  ## where beta is the inverse-scale
   ## gamma(a,b) ~  N (a/b, sd= sqrt(a)/b)
 ## normal approx to Pois
   ## Pois(a) ,  mean =a, var= a
    ## Pois(a) ~ N(a,sd= sqrt(a))
 
 ## gamma prior of total count a-1,  in b observations 
 ## 
 ## we approximate a prior to have an average of 25 deaths across the decade.  the empirical mean is 23.8 ~ 24
 #  the beta parameter is set rate is set such that gamma(25,1) has quantiles  16.2, 35.7 which matches well the observed data.
   alpha= 25
   beta<- 1
   ## the mean rate given the data is 0.126, so we find the beta parameter to match the mean rate.
   n<-nrow(data)

   
  
 plot(theta,dgamma(theta,alpha,rate=beta),lty=2)
lines(theta,dgamma(theta,alpha+n*mean(fatal),rate=beta+n))
 ## credible interval of the posterior
 pois_credible<-qgamma(c(.025,0.975),alpha+n*mean(fatal),rate=beta+n)
 message("posterior gamma interval n=10: ",pois_credible[1]," ",pois_credible[2]) ## 

 post_pois<-dgamma(theta,alpha+n*mean(fatal),rate=beta+n)
  ## posterior mean is 21.9
 sum(theta*post_pois/sum(post_pois))
 
 
 # normal approximation
  mu_0<- alpha/beta  ## average of 12.6 fatal accidents
  tau2_0<- alpha/beta^2 ## 
  
  sigma2<-10^2  ## we assume we know the variance of the fatalities (empirical variance is 22.2) so we set it to 50 (large enough)
  
  
  mu_n<-function(mu0,ybar,n,tau02,sigma2){
  mun<- (mu0/tau02 + (ybar*n)/sigma2)/(1/tau02 + n/sigma2)
  return(mun)
}
taun2<-function(tau02,n,sigma2){
  inv.taun2<- 1/tau02 +n/sigma2
  return(1/inv.taun2)
}

ybar=mean(fatal)

 mn<-mu_n(mu_0,ybar,tau2_0,n,sigma2)
 tn<-taun2(tau2_0,n,sigma2)
## posterior interval n=10
  #lower<-round(qnorm(0.025,mu_n(mu0,ybar,tau02,10,sigma2),sd=sqrt(taun2(tau02,10,sigma2))),2)
  upper<-round(qnorm(c(0.025,0.975),mn,sd=sqrt(tn)),2)

message("posterior interval (normal approx) n=10: ",upper[1]," ",upper[2]) ## 
## the normal approximation is close to the gamma posterior!  woot.
  post_norm<-(dnorm(theta,mn,sqrt(tn))/sum(dnorm(theta,mn,sqrt(tn))))


#### check with simulations ?  check posterior with simulations  posterior mean/sd ?
post.draws<-sample(theta,size=1000,prob=post_pois,replace=T)
 message("the posterior mean is:",mean(post.draws)) ### posterior mean is arout 21.92




   ## check this with the grid approach in the Gamma framework.  
  theta_samps<-sample(theta,100000,prob=post_pois,replace=T)
   ytilde<-rpois(100000,lambda=theta_samps) 
 ## probability of the samples 
 #prob_samp<-post[match(samps,theta)]
  summary(ytilde) ## we have a wide distribution of predictive values.
  
## for all predictive values,  find the total probability  
   ## int p(x|theta)*p(theta|y) d\theta  note that p(x|theta) is a function of theta, so we input hte theta grid.
 pred.prob<-sapply(ytilde,function(x) sum(dpois(x,lambda = theta)*post_pois/sum(post_pois)))  
  
 plot(ytilde,pred.prob, main='predictive distribution for future 1986')
 abline(v=quantile(ytilde,c(0.025,0.975)),col='red')
 lines(theta,post_pois)
 quantile(ytilde,c(0.025,0.975))
```


Using the normal approximation, the variance of the predictive posterior is $\sigma^2+\tau^2$ and is a larger interval (3.9, 44.5).   sampling values from the predictive posterior is much larger interval, but the predictive posterior mean is $\mu_n \approx 24$.  The Gamma conjugate prior has a smaller predictive interval but similar posterior.
```{r}
 
  ## Predictive with normal

## using the normal approximation
## predictive interval must use a grid approach.
     pred<-round(qnorm(c(0.025,0.975),mn,sd=sqrt(tn+sigma2)),2)
  message("predictive interval n=10: ",pred[1]," ",pred[2])

  
   ## check this with the grid approach in the normal framework.  
  post_pred_norm<-(dnorm(theta,mn,sqrt(tn+sigma2))/sum(dnorm(theta,mn,sqrt(tn+sigma2))))
   ytilde_norm<-rnorm(100000,mean=mn,sd=sqrt(tn+sigma2)) 
 ## probability of the samples 
 #prob_samp<-post[match(samps,theta)]
  summary(ytilde_norm) ## we have a wide distribution of predictive values.
  
## for all predictive values,  find the total probability  
   ## int p(x|theta)*p(theta|y) d\theta  note that p(x|theta) is a function of theta, so we input hte theta grid.
 pred.prob_norm<-sapply(ytilde_norm,function(x) sum(dnorm(x,mn,sqrt(tn+sigma2))*post_pred_norm))  
  
 plot(ytilde_norm,pred.prob_norm, main='predictive distribution for future 1986')
 abline(v=quantile(ytilde_norm,c(0.025,0.975)),col='red')
 lines(theta,post_pred_norm)
 quantile(ytilde_norm,c(0.025,0.975))
 
 
 
```

- (b) Assume fatal accidents follow a Poisson distribution with a constant rate and an exposure in each year proportional to the number of passanger miles flown.  Set a prior and determine a posterior and predictive distribution.  estimate the number of passanger miles flown in each year by dividing

let $\theta$ be the fatal accidents per 100,000 million miles flow.  and let $x_i$ be the number per 100,000 million miles flown.   We compute the miles by dividing the passenger deaths divided by the death rate (deaths per 100,000 million miles flown).

We then compute the fatal accident rate by computing the (passanger deaths per 100,000 million miles)*(number of fatal accidents / number of passenger deaths).  We let $\theta$ = fatal accidents per 100 million miles flown.

THe mean accident rate is 0.004, so we set alpha =5 and beta = 1000,  so the prior mean is 0.005 (95$\%$: 0.002,0.01) which is similar to the observed ranges.  

If we take the posterior fatal accident rate per 100,000 miles and multiple by the average miles (5716),  we have the observed fatal accidents as 24.4 (which is approximately 22). The posterior credible interval for average miles per 100,000 is (20.97, 26.97) which is close to the gamma interval of counts (21.1, 27).

For 8 x 10$^11$ miles flown, the fatal accidents is 34.16.


The predictive values were sampled from the posterior distribution, and a random variable was then sampled from a Poisson process with rates equal to the posterior.
```{r}
 miles<-death/rate ## 100 million miles
# print("summary of miles driv")
#summary(miles)
 death_by_accident<-death/fatal
 
 ## fatal accident per 100,000 million miles driven
 acc_rate<-rate*(1/death_by_accident) ## fatal accidents by 100 million miles.
 
 # miles*acc_rate  ## number of fatal accidents.
 print("fatal accident rate summary")
   summary(acc_rate)
  ## crude rate 23.8 fatal accidents by 100 million miles driven.
  
## y ~ Pois( miles*theta)
  ##posterior
 theta=seq(from=0,to=0.1,by=.001)
  alpha= 5
   beta<- 1000  
   ## the mean is 0.005 and quantiles 0.002 and 0.01 which is close to the observed rates 
   #mean(acc_rate) ## fatal accidents per 100,000 million miles driven
   print("prior 95% quantiles: ")
   qgamma(c(0.025,0.975),alpha,rate=beta)
   ## the mean rate given the data is 0.126, so we find the beta parameter to match the mean rate.
   n<-nrow(data)
   
     
  
 plot(theta,dgamma(theta,alpha,rate=beta),lty=2, xlab='fatal accident rate per 100 million miles')
lines(theta,dgamma(theta,alpha+sum(fatal),rate=beta+sum(miles)),col='red')
 ## credible interval of the posterior
 pois_credible<-qgamma(c(.025,0.975),alpha+sum(fatal),rate=beta+sum(miles))
 message("posterior gamma interval n=10: ",round(pois_credible[1],3)," ",round(pois_credible[2],3)) ## 

 
  post_pois<-dgamma(theta,alpha+sum(fatal),rate=beta+sum(miles))

  post.draws<-sample(theta,size=1000,prob=post_pois,replace=T)
   message('posterior mean is: ', mean(post.draws)) ### posterior mean is arout 21.92



   ## check this with the grid approach in the Gamma framework.  
  theta_samps<-sample(theta,100000,prob=post_pois,replace=T)
   ytilde<-rpois(100000,lambda=theta_samps) 
   
 message(" 8,000 per 100 million miles flown has fatality rate: ", mean(ytilde)*8000)
 message("5716 per 100 million miles flown has fatality rate: ", mean(ytilde)*mean(miles))


```
- (c) we repeat part (a) but for passenger deaths.  we need to find a prior and determine the posterior distribution with a predictive interval.

 Note that the death rate mean is not equal to the variance so the performance is not as good compared to the fatal accidents.  The 1986 prediction is 692 (638, 750) which is not the expected predction of 546 passenger deaths.  This is likely because Poisson is not a good fit because it is constrained.
```{r}

 theta=seq(from=200,to=1500,by=.1)

 ## normal approximation to gamma
  ## Gamma(a,b) ,  mean = a/b,  var = a/b^2  ## where beta is the inverse-scale
   ## gamma(a,b) ~  N (a/b, sd= sqrt(a)/b)
 ## normal approx to Pois
   ## Pois(a) ,  mean =a, var= a
    ## Pois(a) ~ N(a,sd= sqrt(a))
 
 ## gamma prior of total count a-1,  in b observations 
 ## 
 ## we approximate a prior to have an average of 600 passenger deaths across the decade.  the empirical mean is 691 
 #  the beta parameter is set rate is set such that   404, 832 which matches well the observed data.
   alpha= 8
   beta<- 0.011
   ## the pr
   n<-nrow(data)

   
  
 plot(theta,dgamma(theta,alpha,rate=beta),lty=2)
lines(theta,dgamma(theta,alpha+n*mean(death),rate=beta+n))
 ## credible interval of the posterior
 pois_credible<-qgamma(c(.025,0.975),alpha+n*mean(death),rate=beta+n)
 message("posterior gamma interval n=10: ",pois_credible[1]," ",pois_credible[2]) ## 

 post_pois<-dgamma(theta,alpha+n*mean(death),rate=beta+n)
  ## posterior mean is 21.9
 sum(theta*post_pois/sum(post_pois))
 
 
 
  ## check this with the grid approach in the Gamma framework.  
  theta_samps<-sample(theta,1000,prob=post_pois,replace=T)
   ytilde<-rpois(1000,lambda=theta_samps) 
 ## probability of the samples 
 #prob_samp<-post[match(samps,theta)]
  summary(ytilde) ## we have a wide distribution of predictive values.
  
## for all predictive values,  find the total probability  
   ## int p(x|theta)*p(theta|y) d\theta  note that p(x|theta) is a function of theta, so we input hte theta grid.
 pred.prob<-sapply(ytilde,function(x) sum(dpois(x,lambda = theta)*post_pois/sum(post_pois)))  
  
 plot(ytilde,pred.prob, main='predictive distribution for future 1986')
 abline(v=quantile(ytilde,c(0.025,0.975)),col='red')
 lines(theta,post_pois)
 quantile(ytilde,c(0.025,0.975))

```
