# Single parameter models {#cross}

## Estimating a probability from binomial data

\begin{equation}
p(y  | \theta) = {n \choose y}\theta^y(1-\theta)^{n-y}
(\#eq:binomialProb)
\end{equation}

To perform Bayesian inference we assume $\theta \sim U(0,1)$ where the posterior is 

\begin{equation}
p(\theta | y) \propto \theta^y(1-\theta)^{n-y}
(\#eq:binomialPosterior)
\end{equation}

which is the form of a `beta` distribution $\theta | y \sim Beta(y+1, n-y+1)$

## Posterior as a compromise between data and prior information
 The posterior is less variable than the prior because it incorporates the information from the data.
 
\begin{equation}
E(\theta) = E(E(\theta | y))
(\#eq:adamsLaw)
\end{equation}

\begin{equation}
V(\theta) = E(V(\theta|y)) + V(E(\theta | y))
(\#eq:evesLaw)
\end{equation}
 
where $\theta|y$ is the posterior.  So the average of the prior, is the average of the posterior means over the distribution of possible data.  The variance of the prior \@ref(eq:evesLaw) says the posterior variance is on average smaller than the prior variance.

## Summarizing the posterior inference
The mean, median, mode, and standard deviation of the posterior distribution summarize the all the current information about a model.  

### Posterior quantiles and intervals{-}

The posterior uncertainty can be reported by presenting the quantiles of the posterior distribution.  The interval, a *central interval of posterior probability* corresponds to the case of 100($1-\alpha)\%$, to the range of values above and below which lies exactly 100($\alpha/2)\%$ of the posterior probability.  The interval estimates are *posterior intervals*.  This differences from the confidence interval because the confidence interval is not a probability interval, because either the parameter is within the region or it is not, but the confidence interval provides information in the long run over repeated experimentation as to how many experiments would contain the true parameter.

There is also the *highest posterior interval* which is a probabilistic interval that is not less than any region outside of the interval.

## Informative prior distributions

the property that the posterior distribution follows the same parametric form as the prior distribution is called *conjugacy*.  Where the beta prior distribution is a *conjugate family* for the binomial likelihood.  

so given the binomial likelihood $p(y|\theta)\propto \theta^a(1-\theta)^b$, and a prior density $p(\theta)\propto \theta^{\alpha-1}(1-\theta)^{\beta-1}$ the posterior is of the beta family.


$$
\begin{aligned}
 p(\theta | y) &\propto \theta^y(1-\theta)^{n-y}\theta^{\alpha-1}(1-\theta)^{\beta-1}\\
 &= \theta^{y+\alpha-1}(1-\theta)^{n-y+\beta-1}\\
 &= Beta(\theta | \alpha+y, \beta+n-y)
\end{aligned}
$$


### Conjugate prior distributions{-}
Conjugacy is formally defined as if F is a class of sampling distributions $p(y|\theta)$, and P is a class of prior distributions for $\theta$, then the class P is conjugate for F if $p(\theta | y) \in P$ for all $p(.|\theta)\in F$ and $p(.)\in P$.

### Conjugate prior, distributions, exponential families, and sufficient statistics{-}
Posterior distributions can be derived using sufficient statistics from exponential families.  The exponential family is defined as 

$$
\begin{aligned}
 p(y_i | \theta)=f(y_i)g(\theta)e^{\phi(\theta)^Tu(y_i)}
\end{aligned}
$$
Where $\phi(\theta), u(y_i)$ are vectors of equal dimension to that of $\theta$.  The $\phi(\theta)$ is called the *natural parameter* for the family (F).  The likelihood of a sequence $y=(y_1,...,y_n)$ iid is

$$
\begin{aligned}
 p(y | \theta)&= (\prod_{i=1}^n f(y_i)) g(\theta)^n exp(\phi(\theta)^T \sum_{i=1}^n u(y_i)) \\
  &\propto g(\theta)^ne^{\phi(\theta)^Tt(y)}, t(y)=\sum_{y=1}^n u(y_i)
\end{aligned}
$$

The *sufficient statistic* for $\theta$ is $t(y)$ because the likelihood for $\theta depends on the data, y, only through the value of t(y).

Sufficient statistics benefit posterior distributions because if the prior density is specified as 
$$
\begin{aligned}
 p(\theta)&\propto g(\theta)^{\eta} e^{\phi(\theta)^T \nu} \\
 \end{aligned}
$$
Then the posterior density using sufficient statistics is

$$
\begin{aligned}
 p(\theta | y ) &\propto g(\theta)^{\eta+n} e^{\phi(\theta)^T (\nu+t(y))} \\
 \end{aligned}
$$

Exponential families are the only classes of distributions that have natural conjugate prior distributions. 

## Normal distribution with known variance

The normal distribution is foundational to statistical modeling, with the central limit theorem (CLT) allowing for the use of normal likelihood in many statistical problems which can approximate complex likelihoods.  If the normal distribution does not provide a good model fit, finite mixtures of distributions can identify useful solutions.

### Likelihood of one data point{-}

With mean $\theta$ and known variance $\sigma^2$ the sampling distribution of a given point is defined 

$p(y|\theta) = \frac{1}{\sqrt{2*\pi*\sigma^2}}e^{-\frac{1}{2\sigma^2} (y-\theta)^2}$

### Conjugate prior and posterior distributions
The prior has the exponential family form given as $\theta \sim N(\mu_0, \tau_0^2)$


$$
\begin{aligned}
 p(\theta ) &\propto exp( \frac{1}{2\tau_0^2}(\theta-\mu_0)^2) \\
 \end{aligned}
$$

WHere completing the square can find the posterior distribution

$$
\begin{aligned}
 p(\theta | y ) &\propto exp(-1/2(\frac{(y-\theta)^2}{\sigma^2} +\frac{(\theta-\mu_0)^2}{\tau_0^2} )  ) \\
 p(\theta | y ) &\propto exp( \frac{1}{2\tau_1^2}(\theta-\mu_1)^2),  \theta|y \sim N(\mu_1,\tau_1^2)
 \end{aligned}
$$
where $\mu_1 = \frac{1/\tau_0^2 \mu_0 + 1/\sigma^2 y}{1/\tau_0^2 + 1/\sigma^2}$ and $1/\tau_1^2 = 1/\tau_0^2 +1/\sigma^2$

In manipulating the distributions the inverse of the variance is defined as the *precision*.  The posterior precision is equal to the prior precision plus the data precision.  And the posterior mean is a weighted average of the prior mean and the observed value, y, proportional to the total precision.

### Posterior predictive distribution{-}

the posterior predictive distribution of a future observation, x, p(x|y) can be calculated 

$$
\begin{aligned}
p(x|y) &= \int p(x|\theta)p(\theta|y) d\theta
&\propto \int exp(-1/2\sigma^2 (x-\theta)^2) exp(-1/2\tau_1^2 (\theta-\mu_1)^2) d\theta
\end{aligned}
$$
 the future observations, x, does not depend on the past observations y given $\theta$.  


### Normal model with multiple observations{-}

For multiple observations, y, the posterior density is formulated as:

$$
\begin{aligned}
 p(\theta | y ) &\propto p(\theta)p(y|\theta) \\
   &=p(\theta)\prod_i p(y_i | \theta)\\
   &\propto exp(-1/2\tau_0^2 (\theta-\mu_0)^2)\prod_i exp(-1/2\sigma^2 (y_i -\theta)^2)\\
   &\propto exp(-1/2 (1/\tau_0^2(\theta-\mu_0)^2+ 1/\sigma^2\sum_i(y_i-\theta)^2) )\\
 \end{aligned}
$$
Simplfying the algebra shows the posterior depends on y only through the sample mean (sufficient statistic), $\bar{y}$ is the sufficient statistic for $\theta$, and the final model is $\bar{y} | \theta,\sigma^2 \sim N(\theta, \sigma^2/n)$

## Other standard single-parameter models
The binomial model is motivated by counting exchangeable outcomes.  The normal distribution applies to a random variable that is the sum of many exchangeable- or independent terms.  The poisson and exponential distribution applies to number of counts, rates, or waiting times for events modeled as occurring exchangeably in all time intervals, i.e. independently in time, and with a constant rate of occurrence.

### Normal distribution with known mean but unknown variance{-}
For $p(y|\theta,\sigma^2)=N(y|\theta,\sigma^2)$ with known mean and unknown variance the likelihood vector with iid observations follows

$$
\begin{aligned}
 p( y | \sigma^2 ) &\propto \sigma^{-n}exp(-1/2\sigma^2 \sum (y_i - \theta)^2) \\
   &= (\sigma^2)^{-n/2}exp(-n/2\sigma^2 \nu)
   \end{aligned}
$$
The sufficient statistic $\nu = -1/n \sum (y_i-\theta)^2$.  The corresponding conjugate prior density is the inverse-gamma 
$$
\begin{aligned}
p(\sigma^2)\propto (\sigma^2)^{-(\alpha+1)}e^{-\beta/\sigma^2}
\end{aligned}
$$
THe convenient parameterization is as a scaled inverse-$\chi^2$ distribution with scale $\sigma_0^2$ and $\nu_0$ degrees of freedom.  The prior distribution for $\sigma^2$ is taken as $\sigma_0^2\nu_0/X$,  where $X\sim \chi_{\nu_0}^2$ random variable.

The resulting posterior density for $\sigma^2$ is 

$$
\begin{aligned}
p(\sigma^2| y) &\propto p(\sigma^2)p(y|\sigma^2)\\
&\propto \frac{\sigma_0^2}{\sigma^2}^{\nu_0/2+1}exp(\frac{\nu_0\sigma_0^2}{2\sigma^2})((\sigma^2)^{-n/2} exp(-n/2\frac{\nu}{\sigma^2}))\\ 
&\propto (\sigma^2)^{-(n+\nu_0)/(2+1)}exp(\frac{-1}{2\sigma^2}(\nu_0\sigma_0^2+n\nu))\\
&\text{then   } \sigma^2 | y \sim Inv.\chi^2 (\nu_0+n, \frac{\nu_0\sigma_0^2+ n\nu}{\nu_0+n})\\
\end{aligned}
$$
Which is a scaled inverse-$\chi^2$ with scale equal to the degrees-of-freedom weighted average of the prior and data scales,  and the d.f. equal to the sum of the prior and data degrees-of-freedom.


### Poisson model{-}
The Poisson model arises in data in the form of counts; in epidemiology studies disease incidence is modeled using Poisson framework.

the likelihood is
$$
\begin{aligned}
p(y|\theta)&= \prod_i^n \frac{1}{y_i!}\theta^{y_i}e^{-\theta}\\
&=\theta^{t(y)}e^{-n\theta}\\
\end{aligned}
$$
Where the sufficient statistic is t(y)=$\sum y_i$ and the likelihood in exponential family form is written as 

$$
p(y|\theta)\propto e^{-n\theta}e^{t(y)log(\theta)}
$$
where the natural parameter $\phi(\theta)=log(\theta)$ and the natural conjugate prior distribution is 

$$
p(\theta)\propto (e^{-\theta})^{\eta}e^{\nu *log(\theta)}
$$
indexed by hyperparameters ($\eta,\nu$).  We can rewrite the likelihood in the form $\theta^a e^{-b\theta}$ and so the conjugate prior must be in the form of $p(\theta)\propto \theta^Ae^{-B\theta}$ which can be re-written to follow a Gamma density as 
$$
p(\theta)\propto e^{-\beta\theta}\theta^{\alpha-1}
$$
and this follows a Gamma($\alpha,\beta)$ equivalent to prior total count of $\alpha-1$ in $\beta$ prior observations.  The posterior is 
$$
\theta|y \sim Gamma(\alpha+n\bar{y}, \beta+n)
$$

### Negative Binomial distribution{-}

## Exercises

### Question 1{-}
prior Beta(4,4), where a coin is tossed 10 times and heads appears fewer than 3 times.  the exact posterior is Beta(4+y, 4+10-y) for y=0,1,2.  Since we don't know the observed heads, but that $y<3$ we plot the posterior distributions for each possibility.  For 2 heads it is closer to the prior, with posterior mean of 0.33,  which is closest to the prior mean of 1/2.
```{r}
 theta<-seq(from=0,to=1,by=0.01)

 plot(theta,dbeta(theta,4,4),type='l',ylim=c(0,5)) ## prior
 lines(theta,dbeta(theta,4+1,4+10-1),lty=2) ## 1 success
   lines(theta,dbeta(theta,4+2,4+10-2),lty=3,col='red') ## 2 succe
   lines(theta,dbeta(theta,4,4+10),lty=4,col='blue')  ## 0 successes
  legend(0.6, 4, 
         legend=c("prior", "0 H", "1 H", "2 H"),
       col=c("black","blue","black", "red"), lty=c(1,4,3,2), cex=0.8)
```


### Normal approximation example{-}


For female births we have beta(438,544) we use the normal approximation.  This replicates Gelman's Figure 2.3 (a, b) 
```{r}
  theta<-seq(from=0.001,to=1,by=0.01)
  ## example births
   postMean <-function(alpha,beta,y,n){
     return( (alpha+y)/(alpha+beta+n))
   } 
   postVar<-function(alpha,beta,y,n){
     return( ((alpha+y)*(beta+n-y))/( (alpha+beta+n)^2*(alpha+beta+n+1)) )
   }
   sdnorm<-sqrt(postVar(438,544,0,0))
   logitMean<-log( postMean(438,544,0,0)/(1-postMean(438,544,0,0)))
   
  logitTheta<-log(theta/(1-theta))

  par(mfrow=c(1,2))
    plot(theta,dbeta(theta,438,544),type='l',xlim=c(0.35,0.55),main="posterior beta") ## prior
    abline(v=0.446,col='red')
  draws<-rbeta(1000,438,544)
  plot(logitTheta,dnorm(logitTheta,mean =logitMean, sd=sd(log(draws/(1-draws))) ),type='l', xlim=c(-0.5,0.1),main="Normal approx.") ## prior
    abline(v=-0.22,col='red')

```



### Question 3{-}


The prior predictive distributions for the number of 6's in a fair roll, tossed 1,000 times will follow a beta distribution.  Let y be the number of 6's in 1000 rolls of fair die, the probability for a 6 is 1/6, so the number of 6's in this trial is approximatley 167, and 833 failures as the prior prediction.  

We plot the beta distribution of the expected number of heads in 1000 tosses.  

The normal approximation shows the probability of heads in a given 1000  tosses, using a non-informative prior beta(167,833) which has a prior predictive mean of $exp(-1.79)/(1+exp(-1.79)) = 0.143$.  This is not the same for number of success in 1000 tosses.  

We find the probability distribution of a given success and the prior probability predictive interval follows a beta with 95% (0.12,0.17) for hte probability of rolling a 6
```{r}
  ## based on normal approximation sketch the distribution of y
  ## for normal we use the logit transform
  ## n = 1,000
  ## first lets construct a beta distribution.
  ## let the prior be beta(4,4) or even beta(1,1)
par(mfrow=c(1,2))

   plot(theta,dbeta(theta,167,833),type='l',xlim=c(0.1,0.5),main="prior beta") ## prior
  # plot(theta,dbeta(theta,1,5),type='l',xlim=c(0.1,0.5), main='prior beta') 
   abline(v=0.167,col='red') ## 
   #  qbeta(c(0.05,0.25,0.5,0.75,0.95),167,833)
    ## predictive on normal scale.
     draws<-rbeta(1000,167,833)
     logitMean<-log( 167/1000)
  sdd<-sd(log(draws/(1-draws)))

   
  plot(logitTheta,dnorm(logitTheta,mean =logitMean, sd=sdd ),type='l',main="Normal approx.of a total heads", xlim=c(-2,-1.4)) ## prior
    abline(v=logitMean,col='red')
    ### the number of heads.    
  message("prior predictive mean:" ,round( exp(logitMean)/(1+exp(logitMean)),2))

 qx<-(qnorm(c(0.05,0.25,0.5,0.75,0.95),mean =logitMean, sd=sdd ))
 ## this is the probability of landing 6 for theta
 exp(qx)/(1+exp(qx))
## wald 95% prior interval
c( exp(logitMean-1.96*sdd)/(1+exp(logitMean-1.96*sdd)),exp(logitMean+1.96*sdd)/(1+exp(logitMean+1.96*sdd)))
```
### Question 4{-}
We have a mixture of 3 normal distributions, and show the central intervals for 5,25,50,75, and 95$\%$ predictive probabilities.  The question gives $\theta$ as the probability of a 6 on a die, possibly unfair, in 1,000 tosses.  we have $\theta= 1/12,1/6, 1/4$ types of biased die. Using the normal approximation, the predictive prior probability is $\sum_{\theta}p(\theta)p(y| \theta)$ where the likelihood is approximated using a normal distribution $\mu = n*\theta_i, \sigma = n*\theta_i(1-\theta_i)$
```{r}
x<-seq(0,1000)-0.5 # continuity correction.
theta<-c(1/12,1/6,1/4)
n=1000
normLikelihood<-function(x,n,p){
  mean<-n*p
  varz<-n*p*(1-p)
  return(dnorm(x,mean=mean,sd=sqrt(varz)))
  
}
a<-dnorm(x,mean=n*theta[1],sd=sqrt(n*theta[1]*(1-theta[1]))  )
b<-dnorm(x,mean=n*theta[2],sd=sqrt(n*theta[2]*(1-theta[2]))  )
c<-dnorm(x,mean=n*theta[3],sd=sqrt(n*theta[3]*(1-theta[3]))  )


mypri<-a*0.25+b*0.5+c*0.25

sum(mypri) ## sums to 1 it is a distribution

 par(mfrow=c(3,2))
plot(x,mypri,type='l',main='predictive prior number of heads')



data<-data.frame(x=x,p=mypri)

## highest probability interval
# 95%

plot(x,mypri,type='l',main=' 95% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.025),1])
)
abline(v=max(data[which(1-cumsum(data$p)>0.025),1]))

plot(x,mypri,type='l',main='75% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.125),1]))
abline(v=max(data[which(1-cumsum(data$p)>0.125),1])
)

plot(x,mypri,type='l',main='50% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.25),1]))
abline(v=max(data[which(1-cumsum(data$p)>0.25),1])
)

plot(x,mypri,type='l',main='25% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.375),1]))
abline(v=max(data[which(1-cumsum(data$p)>0.375),1])
)

plot(x,mypri,type='l',main='5% predictive prior number of heads')
abline(v=max(data[which(cumsum(data$p)<0.475),1]))
abline(v=max(data[which(1-cumsum(data$p)>0.475),1])
)
```

### Question 7{-}
- (a) for the binomial likelihood $y\sim Bin(n,\theta)$, show that p($\theta)\propto \theta^-1(1-\theta)^-1$ is the uniform prior for the natural parameter of the exponential family.
- first we write the binomial likelihood in exponential family form, note that we are not deriving the posterior of any function, but just studying the likelihood.


$$
\begin{aligned}
p(y|\theta) &= \theta^y(1-\theta)^{n-y}\\
\implies log(p)&= ylog(\theta/(1-\theta))+nlog(1-\theta) \\
\implies p(y|\theta) &= exp(y*log(\theta/(1-\theta)))*g(\theta)\\
\text{hence } \phi(\theta)= log(\theta/1-\theta), \text{  and we're given uniformity  } p(\phi(\theta))\propto 1\\
\end{aligned}
$$
Inverting $\phi(\theta)=log(\theta/1-\theta)$ yields $\theta=\frac{e^\phi}{1+e^\phi}$ and we know that p($\phi)\propto 1$.  Using the jacobian we can derive $p(\theta)=p(\phi)\frac{d\phi}{d\theta}$ shown as

$$
\begin{aligned}
p(\theta)&=p(\phi)\frac{d\phi}{d\theta}\\
&= p(\phi)d/d\theta (log(\theta/1-\theta))\\
&\propto (1) (1-\theta / \theta)(1/(1-\theta)^2) \\
&= 1/\theta *(1/1-\theta) 
\end{aligned}
$$

### Question 8 (Normal distribution with unknown mean){-}
A random sample of n students is drawn from a large population, and weights are measured.  The average height of the n sampled students is $\bar{y}= 150$ lbs.  Assume the weights in the population are normally distribution with unknown mean, $\theta$, and known standard deviation 20 lbs.  Suppose the prior for $\theta \sim N(180, 40^2)$

- (a) For known variance, the limit of the posterior is $p(\theta,y)\approx N(\theta| \bar{y}, \sigma^2/n)$.  And the direct formulation is $p(\theta | \bar{y})= N(\theta | \mu_n,\tau_n^2).$ using equations 2.12.

- (b) For a posterior predictive interval, the marginal distribution for new data $p(\tilde{y}  | y) \sim N(\mu_n, \sigma^2+\tau_n^2)$
```{r}
mu_n<-function(mu0,ybar,n,tau02,sigma2){
  mun<- (mu0/tau02 + (ybar*n)/sigma2)/(1/tau02 + n/sigma2)
  return(mun)
}
taun2<-function(tau02,n,sigma2){
  inv.taun2<- 1/tau02 +n/sigma2
  return(1/inv.taun2)
}

```

- (c) Here we give the posterior interval and predictive interval for n=10
```{r}
ybar=150
sigma2=20^2
mu0=180
  tau02=40^2
  n=10

## posterior interval n=10
  #lower<-round(qnorm(0.025,mu_n(mu0,ybar,tau02,10,sigma2),sd=sqrt(taun2(tau02,10,sigma2))),2)
    upper<-round(qnorm(c(0.025,0.975),mu_n(mu0,ybar,tau02,10,sigma2),sd=sqrt(taun2(tau02,10,sigma2))),2)

message("posterior interval n=10: ",upper[1]," ",upper[2])

## posterior predictive interval
  upper2<-round(qnorm(c(0.025,0.975),mu_n(mu0,ybar,tau02,10,sigma2),sd=sqrt(taun2(tau02,10,sigma2)+sigma2) ),2)

message("posterior predictive interval n=10: ",upper2[1]," ",upper2[2])
```

- (d) the posterior and predictive for n=100
```{r}

 lower<-round(qnorm(0.025,mu_n(mu0,ybar,tau02,100,sigma2),sd=sqrt(taun2(tau02,100,sigma2))),2)
    upper<-round(qnorm(0.975,mu_n(mu0,ybar,tau02,100,sigma2),sd=sqrt(taun2(tau02,100,sigma2))),2)

message("posterior interval n=100: ",lower," ",upper)
 ## this approximately equals the limit
print("The asymptotic approximation")
 qnorm(c(0.025,0.975),mean=150, sigma2/100)
## posterior predictive interval
 lower2<-round(qnorm(0.025,mu_n(mu0,ybar,tau02,100,sigma2),sd=sqrt(taun2(tau02,100,sigma2)+sigma2) ),2)
    upper2<-round(qnorm(0.975,mu_n(mu0,ybar,tau02,100,sigma2),sd=sqrt(taun2(tau02,100,sigma2)+sigma2) ),2)

message("posterior predictive interval n=100: ",lower2," ",upper2)
```

### Question 10{-}
Suppose there are N cable cars numbered sequentially 1 to N.  You see a car at random labeled 203 and wish to estimate N.  (a) assume the prior follows Geo(1/100) what is the posterior for N?

- (a) The car X=203 was observed, and the likelihood of this observation (data) is uniform p($X|N=203)=1/N$ which assumes each car is equally likely.  So the posterior $p(N|X) = (1/N)(1/100)(99/100)^{N-1}$.   So the posterior is proportional to $(1/N)(99/100)^{N-1}$ with $N\geq 203$.  

- what is the posterior mean and std. deviation for N?.
   we use Bayes' theorem $p(N|X) = \frac{p(X|N)p(N)}{p(X)}$.  And use a computer to approximation p(X) which is approximately 0.0471 (ignoring constant).
  the infinite series $\sum_{n=203}^\infty (1/n)(99/100)^{n-1}$ converges by using the ratio test with $\rho<1$ to $\approx 0.0471.
  The posterior mean and standard deviation is 279.09 a.nd 79.96, which was evaluated using the posterior distribution $p(N|X)$ defined as
$$
 p(N|X) \propto \frac{(1/N)(99/100)^{N-1}}{p(X)}  
$$
```{r}
 totalSum=0
 
 for(i in 203:30000){
  sum_i= (1/i)*(99/100)^(i-1)
  totalSum= totalSum+sum_i
  
 }
 print(totalSum)
 
 ## approximate the distribution
 px<-totalSum
 X<-seq(203,300000)
 
 dpx<-function(N,px){
   dd<- (1/px)*(1/N)*(99/100)^(N-1)
   
   return(dd)
 }
 ## the posterior mean is
  ## E(X) = x*f(x)
 postMean<-function(N,px){
   fx<- dpx(N,px)
   ex<- N*fx
   return(ex)
 }
  EX<-sum(postMean(X,px))
  
## posterior variance
   ## VX = EX^2 - (EX)^2
postMean.sq<-function(N,px){
   fx<- dpx(N,px)
   ex2<- N*N*fx
   return(ex2)
}
 ## the variance term has to be evaluated at smaller limit to avoid overflow
sdX<-sqrt(sum(postMean.sq(seq(203,35000),px))-(EX)^2)
message("posterior mean and sd is:", round(EX,2)," ",round(sdX,2))
```
- (b) we use a simulation to estimate the posterior.  We sample with replacement, with a distribution size equal to a large number (N), with probability of observing a number equal to the posterior distribution.  The empirial mean and std.dev are 280.5, and 80.9

```{r}
 N<-203:100000
 Nsim<-10000
 unnorm.post<-(1/N)*(99/100)^(N-10)
 mean(sample(N,size=Nsim,prob=unnorm.post,replace=T))
 sd(sample(N,size=Nsim,prob=unnorm.post,replace=T))
```

- (c) we can use a uniform prior $p(N) = 1/N$ with a geometric likelihood p($X|N)=(1/100)(99/100)^(N-1)$ which will not change the result.  We can also change the parameter p to be distributed under a binomial model.
